---
title: "TMA4285 Time series models"
subtitle: "Exercise 8: Comparison of a state space model and a SARIMA model"
author: "Sivert Selnes, Kristine L. Mathisen, Gina Magnussen"
date: "20th of October 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
dataseries <- ts(read.table("dataEx8.txt"))
plot.ts(dataseries)
acf(dataseries)
pacf(dataseries)


```
$\rule[-.3\baselineskip]{0pt}{\baselineskip}$

##### Note how uncertainty should be written: 0.056(7)

# Tasks
## 1. Exploring the data with plotting of relevant statistics
<!--
Do differencing until stationary series (test this, e.g. qq plot) and then find other parameters as before?
Then degree of differencing = d, s = 12 (seen by plot and situation) p,q by acf, pacf and estimation of parameters.
-->

## 2. Justification of the choice of model
## 3. Model parameter estimation including uncertainty 
<!-- Simulation -->
## 4. Model prediction at the given sample points and for the next year including uncertainty of the best linear predictions
<!-- Forecasting -->
## 5. Diagnostics, and model choice discussion including comparison of the two models

# Structure
## Title
## Abstract 
## Introduction
## Theory
### * State-space models
### * SARIMA
## Data analysis
## Discussion
## Conclusion
## Appendix
## References